---
title: "Generating Random Variables"
author: "Stefan Dan,"
date: "2023-05-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Generating U(a,b) from U(0,1)

```{r}
U <- function(n, a, b){
  a + (b-a)*runif(n);
}
print(U(10, 2, 6))
```
Simulation checks
```{r fig.show='hold', out.width="50%"}
a = 4; b = 10;
# prob = T switches y-axis to density
# las = 1 sets the y-axis values vertically
hist(runif(1000, a, b), prob = T, col = "turquoise3", las = 1)
hist(U(1000, a, b), prob = T, col = "green4", las = 1)
```
Inverse transform method

Use the fact that the cdf of $F_x(x)$ is the same as the cdf of $U(0,1)$, i.e 
$F_x(x) \sim U(0,1)$
if $F_x(x)$ is invertible and we can derive the formula analytically, then we 
can generate random variables using

1. Generate $U \sim U(0,1)$
2. Set $X = F^{-1}_x(u)$

More generally, $F^{-1}_x(u) = \inf_x\{F_x(x)=u\}/$

Example: Generate exponentially distributed random variables
$X \sim exp(\lambda)$
pdf: $f(x) = \lambda e^{-\lambda x }$ with $x \in (0, \infty)$
A quick calculation shows
cdf: $F_x(x) = 1 - e^{-\lambda x}$
quantile: $F_x^{-1}(x) = -\ln(1-x)/\lambda$
```{r}
lambda = 2
u = runif(1000)
x = -log(u)/lambda

hist(x, prob = T, las = 1, col="seagreen4",breaks =20,
     main = latex2exp::TeX('$f(x) = \\lambda e^{-\\lambda x }$'))
y = seq(0,4, .01)
lines(y, lambda*exp(-lambda*y), lwd = 2)

```
Acceptance-Rejection Method
```{r}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
