---
title: "Generating Random Variables"
author: "Stefan Dan,"
date: "2023-05-17"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Generating U(a,b) from U(0,1)

```{r}
U <- function(n, a, b){
  a + (b-a)*runif(n);
}
print(U(10, 2, 6))
```
Simulation checks
```{r fig.show='hold', out.width="50%"}
a = 4; b = 10;
# prob = T switches y-axis to density
# las = 1 sets the y-axis values vertically
hist(runif(1000, a, b), prob = T, col = "turquoise3", las = 1)
hist(U(1000, a, b), prob = T, col = "green4", las = 1)
```
Inverse transform method

Use the fact that the cdf of $F_x(x)$ is the same as the cdf of $U(0,1)$, i.e 
$F_x(x) \sim U(0,1)$
if $F_x(x)$ is invertible and we can derive the formula analytically, then we 
can generate random variables using

1. Generate $U \sim U(0,1)$
2. Set $X = F^{-1}_x(u)$

More generally, $F^{-1}_x(u) = \inf_x\{F_x(x)=u\}/$

Example: Generate exponentially distributed random variables
$X \sim exp(\lambda)$
pdf: $f(x) = \lambda e^{-\lambda x }$ with $x \in (0, \infty)$
A quick calculation shows
cdf: $F_x(x) = 1 - e^{-\lambda x}$
quantile: $F_x^{-1}(x) = -\ln(1-x)/\lambda$
```{r}
lambda = 2
u = runif(1000)
x = -log(u)/lambda

hist(x, prob = T, las = 1, col="seagreen4",breaks =20,
     main = latex2exp::TeX('$f(x) = \\lambda e^{-\\lambda x }$'))
y = seq(0,4, .01)
lines(y, lambda*exp(-lambda*y), lwd = 2)

```

$X$ is standard double-exponential distributed i.e
$f(x) = \frac{1}{2}e^{|x|}$
One gets $F_X(x) = \frac{1}{2}e^{x}$ if $x \leq 0$ 

and $F_X(x) = 1-\frac{1}{2}e^{-x}$ if $x>0$

Therefore, $F_X^{-1}(p) = \log(2p)$ if $p\leq 0.5$

and $F_X^{-1}(p) = -\log(2(1-p))$ if $p>0.5$
```{r}
coolerQ <- function(p){
  return(ifelse(p<=0.5, log(2*p), -log(2*(1-p))))
}
n=10000;
u = runif(n);
x = coolerQ(u);
hist(x, prob = T, las = 1, col="yellow4", breaks = 60,
     main = latex2exp::TeX('$f(x) = \\frac{1}{2}e^{|x|}$'),ylim=c(0,0.5), xlim=c(-6,6))
y = seq(-6,6,.02);
lines(y, 0.5*exp(-abs(y)),lwd = 2, col="red")
d <- density(x);
lines(d, lwd =2, col="green")
```

Comparison between empirical and theoretical quantiles

```{r}
p = seq(0.1, 0.9, by = 0.1)
sampleqs = quantile(x,p);
theoryqs = coolerQ(p);
se = sqrt(p*(1-p)/(n*(1/2*exp(-abs(theoryqs)^2))));
devs = round(cbind((sampleqs-theoryqs)/se),2);
df <- data.frame(p,sampleqs,theoryqs,se, devs)
print(df)

```


Acceptance-Rejection Method
```{r}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
